How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex
Added line to table 2
changed program to project

\maketitle
\begin{abstract}
Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education. There is substantial research showing that these visual programming environments are suitable for teaching programming concepts. But, what do people do when they use Scratch? In this paper we explore the characteristics of Scratch programs. To that end we have scraped the Scratch public repository and retrieved 250.000 projects. We present an analysis of these projects in three different dimensions. Firstly, we look at the types of blocks used and the size of the programs. We then investigate complexity, used abstractions and programming concepts. Finally we look at coding style and detect \emph{code smells and clones}. We find that programming abstraction concepts like functions are not commonly used. We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially code duplication.
\end{abstract}

@@

Our study is based on data from the Scratch program repository. By scraping the list of recent programs \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch programs and performed analysis on them.

@@

\begin{itemize}
	\item{A public data set of \nScriptPrograms~Scratch programs (Section \ref{dataset})}
	\item{An evaluation of the data set in terms of (Section \ref{sec:results})}
	\item{A discussion of implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
\end{itemize}

@@

The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs. To answer our three research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch project repository. In the following paragraphs we describe the dataset, the  process and the tools we used for analyzing it, and the methods we followed for detecting code smells.

@@

We obtained the set of Scratch programs by using a scraping program. Our scraping program called Kragle starts by reading \url{https://scratch.mit.edu/explore/projects/all/} and thus obtaining program ids of programs that were most recently shared. Subsequently, Kragle obtains the JSON code for each of the listed programs.

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 programs. Out of the \nPrograms~, we failed to parse and further analyze 2.367 programs due technical difficulties with the provided JSON files. Kragle, as well as all scraped programs and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

Once we obtained the Scratch programs, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per program, with the sprites and the stage of the program. We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks. For example, When Green Flag Clicked is a \emph{Hat block} from the \emph{Events category}.

@@

For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite. For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same program. For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal. We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}. The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar. For this, we examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}. Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.

@@

\subsection{Program Size and Complexity}
\label{RQ1}
The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other programs they were entirety empty apart from the Scratch cat added by default. Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

@@

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrr}
		\hline
		&\textbf{Number of}&\\
		&\textbf{projects}&\textbf{Percentage}\\
		\hline
		Retrieved & 250,166 & \\
		Analyzed & 247,798 & \\
		Non-empty (used for statistics) & \textbf{233,491} & \\
		\hline
		Procedures & 17,979 & 7.70\%\\
		Recursive procedures &  1,052 & 0.45\%\\
		Variables & 73,577 & 31.51\% \\
		Lists & 9,358 & 4.01\% \\
		Conditional statements & 92,959 & 39.81\% \\
		User input blocks& 131,314 & 56.24\% \\
		Loop statements & 180,210 & 77.18\% \\
		\texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
		\texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
		Cloned scripts across sprites & 59,634 & 25.54\% \\
		Cloned scripts within sprites & 23,671 & 10.14\%\\
		Cloned procedures & 4,945 & 2.12\%\\
		Cloned functionality blocks across sprites & 60,554 & 25.93\%\\
		Exact clones across sprites & 27,574 & 11.81\%\\
		Exact clones within sprites & 2,043 & 0.87\%\\
		Dead code & 56,890	& 24.36\%\\
		Large scripts & 69,521 & 29.77\%\\
		Large sprites & 31,954 & 13.68\%\\		
		\hline
	\end{tabular}
	\caption{Characteristics in the projects in the dataset}
	\label{tbl-characteristics}
\end{table*}

@@

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project. We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.

@@

To understand the complexity of the Scratch programs in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

\subsubsection{Dead Code}
In our analysis, we find that almost one quarter of the Scratch programs contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality. \todo{Fenia, this is correct, right, the 23\% is just these types and not also uncalled procs and unmatched broadcasts?}.

@@

\subsubsection{Exact Clones between Sprites}
With occurrences in 11\% of the Scratch programs in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no other way to share the functionality than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

\subsubsection{Sharing of Scripts and Procedures}
Investigating the use of clones between programs, we observe that there are 1700 scripts that are used in multiple programs, sometimes as often as in 1600 different programs. This seems to indicate that there are common patterns in Scratch programs, which means it might be very beneficial to Scratch programmers if they could not only share their programs, but also share some of their functionality, for others to use, like a library. An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'. This might empower new Scratch users to get started quicker.

\subsection{Threads to validity}
A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 programs. It could be the case that programming habits of Scratch users are changing over time. However, we counterbalanced that by using a very large dataset. Furthermore, for some of the analyses we manually inspected programs \todo{Fenia, which ones?}

Furthermore we use the number of blocks in the Scratch programs as a measure for the length of a program, while this does not exactly correspond to the `length' of a program in lines, and there can be multiple Scratch blocks on one line. For example, in Figure \ref{fig:ui} the `if' and `touching' blocks are present on the same `line'. We however believe that the number of blocks is a good proxy for size, and we plan a future experiment in which we will compare `lines of Scratch code' to `number of blocks'.

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that functions and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.

>>>>

\maketitle
\begin{abstract}
Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education. There is substantial research showing that these visual programming environments are suitable for teaching programming concepts. But, what do people do when they use Scratch? In this paper we explore the characteristics of Scratch programs. To that end we have scraped the Scratch public repository and retrieved 250.000 projects. We present an analysis of these projects in three different dimensions. Firstly, we look at the types of blocks used and the size of the projects. We then investigate complexity, used abstractions and programming concepts. Finally we look at coding style and detect \emph{code smells and clones}. We find that programming abstraction concepts like procedures are not commonly used. We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially code duplication.
\end{abstract}

@@

Our study is based on data from the Scratch project repository. By scraping the list of recent projects \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them.

@@

\begin{itemize}
	\item{A public data set of \nScriptPrograms~Scratch projects (Section \ref{dataset})}
	\item{An evaluation of the data set in terms of (Section \ref{sec:results})}
	\item{A discussion of implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
\end{itemize}

@@

The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch projects. To answer our three research questions, we conducted an empirical quantitative evaluation of project data we collected from the Scratch project repository. In the following paragraphs we describe the dataset, the  process and the tools we used for analyzing it, and the methods we followed for detecting code smells.

@@

We obtained the set of Scratch projects by using a scraping program. Our scraping program called Kragle starts by reading \url{https://scratch.mit.edu/explore/projects/all/} and thus obtaining project ids of projects that were most recently shared. Subsequently, Kragle obtains the JSON code for each of the listed projects.

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects. Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project. We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks. For example, When Green Flag Clicked is a \emph{Hat block} from the \emph{Events category}.

@@

For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite. For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project. For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal. We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}. The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar. For this, we examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}. Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.

@@

\subsection{Program Size and Complexity}
\label{RQ1}
The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other projects they were entirety empty apart from the Scratch cat added by default. Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

@@

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrr}
		\hline
		&\textbf{Number of}&\\
		&\textbf{projects}& \textbf{Percentage}\\
		\hline
		Retrieved & 250,166 & \\
		Analyzed & 247,798 & \\
		Non-empty (used for statistics) & \textbf{233,491} & \\

		\textit{Projects with:} & & \\
		\hline
		Procedures & 17,979 & 7.70\%\\
		Recursive procedures &  1,052 & 0.45\%\\
		Variables & 73,577 & 31.51\% \\
		Lists & 9,358 & 4.01\% \\
		Conditional statements & 92,959 & 39.81\% \\
		User input blocks& 131,314 & 56.24\% \\
		Loop statements & 180,210 & 77.18\% \\
		\texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
		\texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
		Cloned scripts across sprites & 59,634 & 25.54\% \\
		Cloned scripts within sprites & 23,671 & 10.14\%\\
		Cloned procedures & 4,945 & 2.12\%\\
		Cloned functionality blocks across sprites & 60,554 & 25.93\%\\
		Exact clones across sprites & 27,574 & 11.81\%\\
		Exact clones within sprites & 2,043 & 0.87\%\\
		Dead code & 56,890	& 24.36\%\\
		Large scripts & 69,521 & 29.77\%\\
		Large sprites & 31,954 & 13.68\%\\		
		\hline
	\end{tabular}
	\caption{Elements and characteristics of the projects in the dataset}
	\label{tbl-characteristics}
\end{table*}

@@

In Table \ref{tbl-size} we summarize statistics for the analyzed metrics. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project. We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.

@@

To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

\subsubsection{Dead Code}
In our analysis, we find that almost one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality. \todo{Fenia, this is correct, right, the 23\% is just these types and not also uncalled procs and unmatched broadcasts?}.

@@

\subsubsection{Exact Clones between Sprites}
With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no other way to share the functionality than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

\subsubsection{Sharing of Scripts and Procedures}
Investigating the use of clones between projects, we observe that there are 1700 scripts that are used in multiple projects, sometimes as often as in 1600 different projects. This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library. An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'. This might empower new Scratch users to get started quicker.

\subsection{Threads to validity}
A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 projects. It could be the case that programming habits of Scratch users are changing over time. However, we counterbalanced that by using a very large dataset. Furthermore, for some of the analyses we manually inspected projects \todo{Fenia, which ones?}

Furthermore we use the number of blocks in the Scratch projects as a measure for the length of a program, while this does not exactly correspond to the `length' of a program in lines, and there can be multiple Scratch blocks on one line. For example, in Figure \ref{fig:ui} the `if' and `touching' blocks are present on the same `line'. We however believe that the number of blocks is a good proxy for size, and we plan a future experiment in which we will compare `lines of Scratch code' to `number of blocks'.

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that procedures and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.
